Enzo: An Adaptive Mesh Refinement Code for Astrophysics

### 2013-06-08 numerical-ppm.tex
More changes to PPM - cleaned up and made more general to be useful for other Godunov methods.

@@

%We can examine the structure of Equations~(\ref{eq:mass})-(\ref{eq:total_energy})
%to obtain a better understanding of the effect of each term.
%The derivatives are first order in time
%and provide us with the dependent variables that will be evaluated each
%timestep: $\rhob$, $\vecvb$, and $E$.
%
%The second term on the left hand side of each relation
%is due to our choice of Eulerian coordinates and corresponds to
%advection, the flow of conserved quantities 
%($\rhob$, $\rhob \vecvb$, and $\rhob E$) 
%across the mesh.  These will be termed the transport terms.

The first term on the right hand side of equations~(\ref{eq:momentum}) and
(\ref{eq:total_energy}) come from our choice of comoving coordinates
(a similar term also appears in equation~(\ref{eq:dm_velocity}), the velocity 
relation for the dark matter particles).  A similar term does not
appear in the mass conservation equation (\ref{eq:mass}) because
of the comoving density definition.   
We note that these expansion terms could be
eliminated entirely by the proper choice of variables (including time),
although we have not done so here, as they do not constitute a major
source of error.  
We solve these terms (for all hydro methods)
using the same technique -- they are split from the rest of the terms and
solved using an (implicit) time-centered method, which is straightforward
as there are no spatial gradients.

%
%The remainder of the right-hand side terms (due to pressure and gravity) are denoted as
%source terms.  We make these distinctions because the 
%source and transport steps will be solved in
%a self-consistent fashion by the PPM scheme while the expansion terms,
%which are spatially localized, are split off and solved in a separate step
%(as are radiative cooling, heating and other processes, if present).

The remainder of the terms in the fluid equations
involve derivatives.  
Because we are interested in phenomena with no special geometry, we will
restrict discussion to Cartesian coordinates.
Once again we use the concept of operator splitting,
but now apply it spatially.   We can rewrite the one-dimensional (Eulerian) 
versions of equations~(\ref{eq:mass})--(\ref{eq:total_energy}) without expansion terms
in conservative form as,
\begin{eqnarray}
\frac{\partial \rho}{\partial t}  + \frac{1}{a} \frac{\partial \rho v }{\partial x}    & = &  
     0 \label{eq:mass1d} \\
\frac{\partial \rho v}{\partial t}  + \frac{1}{a} \frac{\partial \rho v^2}{\partial x}   + 
      \frac{1}{a} \frac{\partial p}{ \partial x} & = & 
      \rho \frac{1}{a} \frac{\partial \phi}{ \partial x}  \label{eq:momentum1d}  \\
\frac{\partial \rho E}{\partial t}  + \frac{1}{a} \frac{\partial \rho v E}{\partial x}  & =  &
      \rho v \frac{1}{a} \frac{\partial \phi}{\partial x}. \label{eq:total_energy1d}
\end{eqnarray}

%
Note that $x$ and $v$ refer to the one dimensional comoving position and 
peculiar velocity of the baryonic gas, and $g$ is the acceleration at the cell center.
We write this including the cosmological scale factors, but it reduces to the non-cosmological case with $a = 1$.
These equations are now in a form that can be solved by the split PPM scheme.

@@

We now restrict ourselves to the solution of equations
(\ref{eq:mass1d})--(\ref{eq:total_energy1d}), in one
dimension.  In the Direct Eulerian version of PPM, this is
accomplished by a three-step process.  First, we compute effective
left and right states at each grid boundary by constructing a
piecewise parabolic description of the primative variables ($\rho$, $u$
and $E$) and then averaging over the regions corresponding to the
distance each characteristic wave can travel ($u$, $u-c_s$ and
$u+c_s$, where $c_s$ is the sound speed).
Second, a Riemann problem is solved using these effective left and
right states, and finally the fluxes are computed based on the
solution to this Riemann problem.  This is described in detail in
\citet{1984JCoPh..54..174C}, but we will briefly outline the procedure
here both for completeness and to put the changes we will make in
context.

The Eulerian difference equations are:
%
\newcommand{\avp}[1]{\overline{#1}_{j+1/2}}
\newcommand{\avm}[1]{\overline{#1}_{j-1/2}}
\begin{equation}
\rho_j^{n+1} = \rho_j^{n} + \Delta t \left( 
                      \frac{ \avp{\rho}\avp{v} -  \avm{\rho}\avm{v} } {\Delta x_j}
            \right)
     \label{eq:mass_diff}
\end{equation}
\begin{equation}
\rho_j^{n+1} v_{j}^{n+1} =
       \rho_j^{n+1} v_j^n  + \Delta t \left(
          \frac{ \avp{\rho} \avp{v}^2 - \avm{\rho} \avm{v}^2 + \avp{p} - \avm{p}} {\Delta x_j}
             \right)
     + \frac{ \Delta }{2} g_j^{n+1/2} (\rho_j^n + \rho_j^{n+1}), 
     \label{eq:momentum_diff}
\end{equation}
\begin{eqnarray}
\rho_j^{n+1} E_j^{n+1}  = 
       \rho_j^n E_j^n  & + & \Delta t  \left(
            \frac{  \avp{\rho} \avp{v} \avp{E}  - \avm{\rho} \avm{v} \avm{E}  +
                       \avp{v} \avp{p}    - \avm{v} \avm{p} } {\Delta x_j}
             \right) \nonumber \\
         & & + \frac{ \Delta t }{2} g_j^{n+1/2} (\rho_j^n v_j^n + \rho_j^{n+1} v_j^{n+1} ).
     \label{eq:energy_diff}
\end{eqnarray}

@@

We first construct monotonic piecewise 
parabolic (third order) interpolations in one dimension
for each of $p$, $\rho$, and $v$.  The pressure is determined from
equation~(\ref{eq:eq_of_state}), the equation of state.
%We have dropped the subscript b in this and the following two sections because we will be referring solely to baryonic quantities.  
The interpolation formula for some quantity $q$ is given by:
%
\begin{eqnarray}
q_j(x) & = &  q_{L,j} + \tilde{x}(\Delta q_j + q_{6,j}(1-\tilde{x})), \\
\tilde{x}      & \equiv & {x - x_{j-1/2} \over \Delta x_j}, \qquad
             x_{j-1/2} \leq x \leq x_{j+1/2}. \nonumber
\end{eqnarray}
%
This is Equation~(1.4) of CW84 (in the spatial rather than mass coordinate as is appropriate for the direct Eulerian approach). 
%The zones edges in the mass coordinates are $m_{j-1/2}$ and $m_{j+1/2}$ for the left and right edges, respectively.  
The quantities $q_{L,j}$, $\Delta q_j$,
and $q_{6,j}$ can be viewed simply as interpolation constants; however,
they also have more intuitive meanings.  For example, $q_{L,j}$ is the
value of $q$ at the left edge of zone j, while $\Delta q_j$ and $q_{6,j}$ are analogous to the slope and first order correction to the slope of $q$ (see \citet{1984JCoPh..54..174C} for a complete discussion):
\begin{equation}
\Delta q_j \equiv q_{R,j} - q_{L,j} \qquad 
q_{6,j}    \equiv 6\left[q_j - 1/2\left(q_{L,j} + q_{R,j}\right)\right].
\end{equation}

@@

Once we have the reconstruction, the primary quantities ($p$, $\rho$, $v$ and $E$) are averaged over the domains corresponding to the three characteristics $C = u-c$, $u$ or $u+c$ (where $c$ is the sound speed in a cell):
\begin{eqnarray}
      q^{C+}_{j+1/2} & = &
          {1 \over \Delta t C^n_j} \int^{x_{j+1/2}}_{x_{j+1/2}-\Delta t C^n_j} q_j(x) dx, \\
      q^{C-}_{j+1/2} & = & 
         {1 \over -\Delta t C^n_{j+1}} \int^{x_{j+1/2}+\Delta t C^n_{j+1}}_{x_{j+1/2}} q_{j+1}(x) dx \nonumber
\end{eqnarray}
where $q^{C+}$ ($q^{C-}$) refers to the left (right) characteristic $C$.  The linearized gas dynamics equations are then used to compute second-order accurate left and right states that take into account the multiple wave families.  This process is described in \citet{1984JCoPh..54..174C} and we use their equations (3.6) and (3.7).

With these effective states, an approximation to the Riemann problem is found
with an iterative approach \citep[see][and section~\ref{sec.riemann}]{Woodward86}, producing estimates for 
$\overline{p}_{j\pm 1/2}$, $\overline{\rho}_{j\pm1/2}$, and $\overline{v}_{j\pm 1/2}$ that are third order
accurate in space and second order accurate in time.  These are then
used to solve the difference Equations (\ref{eq:mass_diff})--(\ref{eq:energy_diff}) for $\rho^{n+1}$, $v^{n+1}$, and $E^{n+1}$.

@@

While most numerical hydrodynamics schemes center around finding approximate
solutions to the hydrodynamics equations, the core philosophy of a Godunov
method \citep{Godunov1959} is to find the exact solution to an
approximation of the physical setup.  The approximate problem in this case is
the Riemann problem, wherein two volumes are separated by a thin membrane that
is instantaneously removed at time $t=0$.  The subsequent evolution has an exact 
analytic solution.  This solution is
described in detail in many texts on computational fluid dynamics
\citep[e.g.,][]{toro-1997}.   In brief, there are three waves that propagate
away from the initial discontinuity.  The central wave, characterized by a
density jump but not a pressure jump, is called the contact discontinuity.  The
waves traveling to the left and right of the contact discontinuity can be either
shocks, if characteristics converge on the wave front, or  rarefaction fans if
characteristics diverge.

@@

On rare occasions, high order
solutions can cause negative densities or energies.  Both PPM and MUSCL solvers employ a Riemann
solver fallback mechanism \citep{Lemaster09}.  If a negative density is found at
a particular interface, the more diffusive HLL Riemann solver is used to compute the fluxes associated with that cell.

>>>>

In this section, we describe the four solvers that we have implemented for
solving the fluid equations.  We describe the PPM method in considerably
more detail than the other methods in part
because its implementation in Enzo has not previously been described,
but mostly because it introduces many of the ideas and methods used for
the MUSCL-Dedner and MHDCT schemes (including expansion terms,
reconstruction, Riemann solvers, and dual energy formalism);

@@

The first term on the right hand side of equations~(\ref{eq:momentum}) and
(\ref{eq:total_energy}) come from our choice of comoving coordinates
(a similar term also appears in equation~(\ref{eq:dm_velocity}), the velocity 
relation for the dark matter particles).  A similar term does not
appear in the mass conservation equation (\ref{eq:mass}) because
of the comoving density definition.   
We note that these expansion terms could be
eliminated entirely by the proper choice of variables (including time),
although we have not done so here, as they do not constitute a major
source of error.  We solve these terms
using the same technique -- they are split from the rest of the terms and
solved using an (implicit) time-centered method, which is straightforward
as there are no spatial gradients.  Note that we use this method for
all four of the hydro solvers.

The remainder of the terms in the fluid equations
involve derivatives.  
Because we are interested in phenomena with no special geometry, we will
restrict discussion to Cartesian coordinates.
We can dimensionally split the equations and rewrite the one-dimensional (Eulerian) 
versions of equations~(\ref{eq:mass})--(\ref{eq:total_energy}) without expansion terms
in conservative form as,
\begin{eqnarray}
\frac{\partial \rho}{\partial t}  + \frac{1}{a} \frac{\partial \rho v }{\partial x}    & = &  
     0 \label{eq:mass1d} \\
\frac{\partial \rho v}{\partial t}  + \frac{1}{a} \frac{\partial \rho v^2}{\partial x}   + 
      \frac{1}{a} \frac{\partial p}{ \partial x} & = & 
      \rho \frac{1}{a} \frac{\partial \phi}{ \partial x}  \label{eq:momentum1d}  \\
\frac{\partial \rho E}{\partial t}  + \frac{1}{a} \frac{\partial \rho v E}{\partial x}  & =  &
      \rho v \frac{1}{a} \frac{\partial \phi}{\partial x}. \label{eq:total_energy1d}
\end{eqnarray}
%
Here $x$ and $v$ refer to the one dimensional comoving position and 
peculiar velocity of the baryonic gas, and $g$ is the acceleration at the cell center.
These equations are now in a form that can be solved by the split PPM scheme.

@@

We now restrict ourselves to the solution of equations
(\ref{eq:mass1d})--(\ref{eq:total_energy1d}), in one
dimension.  In the Direct Eulerian version of PPM, this is
accomplished by a three-step process.  First, we compute effective
left and right states at each grid boundary by constructing a
piecewise parabolic description of the primative variables ($\rho$, $u$
and $E$) and then averaging over the regions corresponding to the
distance each characteristic wave can travel ($u$, $u-c_s$ and
$u+c_s$, where $c_s$ is the sound speed).
Second, a Riemann problem is solved using these effective left and
right states, and finally the fluxes are computed based on the
solution to this Riemann problem and the conserved quantities are
updated.  This is described in detail in
\citet{1984JCoPh..54..174C}, but we will briefly outline the procedure
here both for completeness and to put the changes we will make in
context.

The Eulerian difference equations are:
%
\newcommand{\avp}[1]{\overline{#1}_{j+1/2}}
\newcommand{\avm}[1]{\overline{#1}_{j-1/2}}
\begin{equation}
\rho_j^{n+1} = \rho_j^{n} + \Delta t \left( 
                      \frac{ \avp{\rho}\avp{v} -  \avm{\rho}\avm{v} } {\Delta x_j}
            \right)
     \label{eq:mass_diff}
\end{equation}
\begin{equation}
\rho_j^{n+1} v_{j}^{n+1} =
       \rho_j^{n+1} v_j^n  + \Delta t \left(
          \frac{ \avp{\rho} \avp{v}^2 - \avm{\rho} \avm{v}^2 + \avp{p} - \avm{p}} {\Delta x_j}
             \right)
     + \frac{ \Delta t }{2} g_j^{n+1/2} (\rho_j^n + \rho_j^{n+1}), 
     \label{eq:momentum_diff}
\end{equation}
\begin{eqnarray}
\rho_j^{n+1} E_j^{n+1}  = 
       \rho_j^n E_j^n  & + & \Delta t  \left(
            \frac{  \avp{\rho} \avp{v} \avp{E}  - \avm{\rho} \avm{v} \avm{E}  +
                       \avp{v} \avp{p}    - \avm{v} \avm{p} } {\Delta x_j}
             \right) \nonumber \\
         & & + \frac{ \Delta t }{2} g_j^{n+1/2} (\rho_j^n v_j^n + \rho_j^{n+1} v_j^{n+1} ).
     \label{eq:energy_diff}
\end{eqnarray}

@@

We first construct monotonic piecewise 
parabolic (third order) interpolations in one dimension
for each of $p$, $\rho$, and $v$.  The pressure is determined from
equation~(\ref{eq:eq_of_state}), the equation of state.
The interpolation formula for some quantity $q$ is given by:
%
\begin{eqnarray}
q_j(x) & = &  q_{L,j} + \tilde{x}(\Delta q_j + q_{6,j}(1-\tilde{x})), \\
\tilde{x}      & \equiv & {x - x_{j-1/2} \over \Delta x_j}, \qquad
             x_{j-1/2} \leq x \leq x_{j+1/2}. \nonumber
\end{eqnarray}
%
This is equation~(1.4) of CW84 (in the spatial rather than mass coordinate as is appropriate for the direct Eulerian approach). 
The quantities $q_{L,j}$, $\Delta q_j$,
and $q_{6,j}$ can be viewed simply as interpolation constants; however,
they also have more intuitive meanings.  For example, $q_{L,j}$ is the
value of $q$ at the left edge of zone j, while $\Delta q_j$ and $q_{6,j}$ are analogous to the slope and first order correction to the slope of $q$ (see \citet{1984JCoPh..54..174C} for a complete discussion):
\begin{equation}
\Delta q_j \equiv q_{R,j} - q_{L,j} \qquad 
q_{6,j}    \equiv 6\left[q_j - 1/2\left(q_{L,j} + q_{R,j}\right)\right].
\end{equation}

@@

Once we have the reconstruction, the primary quantities ($p$, $\rho$, $v$ and $E$) are averaged over the domains corresponding to the three characteristics $C = u-c$, $u$ or $u+c$ (where $c$ is the sound speed in a cell):
\begin{equation}
      q^{C+}_{j+1/2} =
          {1 \over \Delta t C^n_j} \int^{x_{j+1/2}}_{x_{j+1/2}-\Delta t C^n_j} q_j(x) dx
\end{equation}
where $q^{C+}$ refers to the left characteristic $C$ (and a similar integral is done for the right characteristic $q^{C-}$).  The linearized gas dynamics equations are then used to compute second-order accurate left and right states that take into account the multiple wave families.  This process is described in \citet{1984JCoPh..54..174C} and we use their equations (3.6) and (3.7).

With these effective states, an approximation to the Riemann problem is found
(see below for more detail about the Riemann solvers used), producing estimates for 
$\overline{p}_{j\pm 1/2}$, $\overline{\rho}_{j\pm1/2}$, and $\overline{v}_{j\pm 1/2}$ that are third order
accurate in space and second order accurate in time.  These are then
used to solve the difference Equations (\ref{eq:mass_diff})--(\ref{eq:energy_diff}) for $\rho^{n+1}$, $v^{n+1}$, and $E^{n+1}$.

@@

In this section, we describe the methods we adopt to solving the Riemann
problem, which is generally required to compute the fluxes in 
any Godunov.  This section therefore applies to all three of our 
Godunov-based schemes.  The Riemann problem we are solving
involves two constant states separated by a single discontinuity at $t=0$.
The subsequent evolution has an exact analytic solution.  This solution is
described in detail in many texts on computational fluid dynamics
\citep[e.g.,][]{toro-1997}.   In brief, there are three waves that propagate
away from the initial discontinuity.  The central wave, characterized by a
density jump but not a pressure jump, is called the contact discontinuity.  The
waves traveling to the left and right of the contact discontinuity can be either
shocks, if characteristics converge on the wave front, or rarefaction fans if
characteristics diverge.

@@

On rare occasions, high order solutions can cause negative densities or energies.  
Both our PPM and MUSCL solvers employ a Riemann
solver fallback mechanism \citep{Lemaster09}.  If a negative density is found at
a particular interface, the more diffusive HLL Riemann solver is used to compute the fluxes associated with that cell, and the flux update is repeated.
